### 2024-04-03 doc/15yrsChange.Rmd
Results and discussion, misc. edits

Erika Gonzalez-Akre^1^

@@

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, and successional trajectories. [@lovett_nonnative_2016; @herms_emerald_2014]. Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.

@@

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021]

@@

We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* [@gonzalez-akre_allodb_2022]. The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information. Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot study regions and the entire 25.6-ha plot. We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region. We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data.

In addition to the estimation of AWR biomass flux, we analyzed the effects of nuisance species on recruitment through counts of newly recruited stems. We categorized each species as either canopy or understory based on expert knowledge of life history traits. *Quercus* spp. and *Carya* spp. were grouped by genera to better assess differences between the recruitment of canopy and understory species. We then calculated the number of new stems for each census interval for each species or genera. Finally, we assessed the spatial overlap of biomass loss and recruitment of canopy species. We calculated the net change in AGB within each 0.4 hectare quadrat  between 2008 and 2023 and the AGB of canopy species in the understory in 2023.

@@

For the first time since the establishment of  the SCBI ForestGEO plot, whole-plot aboveground C stocks decreased over the 2018-2023 census interval (Figure 2) . From 2018 to 2023,ABG decreased from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^. Prior to this, ABG had steadily increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a). In regions with  high deer density and high canopy vulnerability AGB decreased significantly over the past two censuses: from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013, to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018, and `r hdhn_AGB_2023` Mg C ha^-1^ in 2023. Areas with low density of vulnerable canopy species experienced small losses of AGB between 2018 and 2023; inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.

![**_Figure `r biomass_trends`. Aboveground biomass at the SCBI ForestGEO plot since 2008 establishment._** Values are shown for the plot as a whole and for habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure2.jpeg)

The trend in net woody production can be explained by its constituent fluxes (Figure 3). Trends in woody mortality show that for the first two census intervals, the whole plot has lost **VALUE** and then **VALUE** Mg C/ha/year, increasing to a loss of **VALUE** Mg C/ha/year for the most recent census interval (Figure 3b). Woody productivity contributes positively to the flux, showing that the whole plot gained **VALUE** Mg C/ha/year in the first interval, **VALUE** Mg C/ha/year in the second, and **VALUE** Mg C/ha/year in the third (Figure 3c). Areas of low deer, low nuisance species were consistently the most productive ending at **VALUE** Mg C/ha/year in the interval 2018 to 2023, and areas of high deer, high nuisance species were the least productive at **VALUE** Mg C/ha/year in the interval 2018 to 2023. Recruitment has consistently added smaller magnitude amounts of carbon with the whole plot gaining **VALUE** Mg C/ha/year (Figure 3d). Recruitment was highest at **VALUE** Mg C/ha/year in areas with high deer and high vulnerable species, and lowest at **VALUE** Mg C/ha/year in areas with high deer and low vulnerable species.

![**_Figure `r loss_gain_net`. Changes in aboveground woody biomass at the SCBI ForestGEO plot over each census interval since 2008 establishment, including (a) net change, (b) losses to mortality, and gains from (c) woody productivity and (d) recruitment._** Values are shown for the plot as a whole and for habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure3.jpeg)

@@

Woody mortality was the primary driver of the negative trend in biomass, and certain species were responsible for higher levels of mortality. Fraxinus americana had the highest aboveground woody mortality rates of any species present on the plot, peaking at **VALUE** Mg C/ha/year lost in *2020* , and **VALUE** Mg C/ha/year lost in 2022 (Figure 4). Quercus velutina contributed the second-highest levels of mortality, losing **VALUE** Mg C/ha/year in 2022. Quercus rubra, Carya species, Quercus prinus, Quercus alba, and Liriodendron tulipifera were all significant contributors as well.

![**_Figure `r mortality_by_genera`. Annual woody mortality by species at the SCBI ForestGEO plot from 2008-2023._** Shown are the seven canopy taxa with highest average mortality rate in descending order and a category including all others. The census included all stems ≥ 10cm in the most recent full census.](display/Figure4.jpeg)

## Species composition of recruitment flux
Certain species dominated aboveground woody recruitment, with understory species being recruited at higher rates than canopy species (Figure 5). Out of the canopy species, Carya species showed the highest levels of recruitment with a rate of **VALUE** Mg C/ha/year recruited from 2018 to 2023 (Figure 5a). Acer negundo had the second highest rate of recruitment for canopy species at **VALUE** Mg C/ha/year from 2018 to 2023. Fagus grandifolia, Fraxinus americana, and Ulmus rubra are the other species with the highest recruitment rates. Most canopy species had lower recruitment rates in the 2018 to 2023 census interval than the 2008 to 2013 interval. Lindera benzoin had by far the highest recruitment rate of the understory species at **VALUE** Mg C/ha/year from 2018 to 2023 (Figure 3b). Asimina triloba had the second highest rate of recruitment of the understory species at **VALUE** Mg C/ha/year recruited from 2018 to 2023. Other understory species with high recruitment were Carpinus caroliniana, Hamamelis virginiana, and Celtis occidentalis.

![**_Figure `r recruitment_by_genera`. Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment >20x that of canopy species recruitment.](display/Figure5.jpeg)

## Regeneration & mortality mismatch

![**_Figure `r regen_failure`._**](display/Figure6.jpeg)

@@

We found that the combination of decreasing biomass (driven by increasing woody mortality) and limited recruitment of canopy species in the understory is compromising the potential the carbon sink, now and into the future. 
Differences in AGB were the biggest in the regions of the plot with high vulnerability to nuisance species. These regions also had the lowest amount of canopy species in the understory. Areas with low deer population had the highest amount of canopy species in the understory.
Some of the plot is increasing biomass, notably those with low proportions of vulnerable species. On net, these increases are not outpacing mortality. 
Suggests that the combination of heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to pathogen-derived mortality where understory species that are resistant to deer and fast growers can take advantage, further competing with canopy species.
While Emerald Ash-borer has had a large impact on biomass loss at our site, mortality among oak species, which is not linked to a non-indigenous pest, is substantial as well. Oak species in the region are generally found in older, even aged cohorts, which contributes to accelerated mortality rates (Oak et al., 2016). However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century. These mortality events are often termed “oak decline” events due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992). Oak decline is believed to be the result of an imbalance in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).

Using current understory as a predictor of the future canopy is imperfect. 
Some canopy species (examples) recruit in pulse events - if there is a shift in deer population and enough survival that coincides with a mast year than you could get regeneration into the canopy.
However, current recruitment levels are below a rate at which we would expect in maturing secondary forests and may be insufficient to prevent large periods without a close canopy.
Our data is from just the single site, which can make generalizability challenging. However, there is a lot of support that the drivers we identify here are representative of the region.
Mixed hardwood forests are typical of mid-Atlantic forests and deer populations are high throughout the region.
Vulnerable species (ash) are also common throughout these forests, and emerald ash borer continues to spread. Where it does spread it consistently decimates adult ash.
Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forests. 
Nuisance species (driven by human pressures/actions) interact with other anthropogenic changes to modify forest ecosystems. 
Have the capacity to alter short-term dynamics: large carbon losses that would not be expected in models that don’t exclude it.
Longer-term have capacity to alter successional dynamics: can delay recovery from disturbances (including other disturbances) or even suppress regeneration into a closed-canopy for sustained periods. 
Decrease resilience to other pressures as pests and pathogens decrease diversity through shrinking the number of possible canopy species.
Positive takeaways: this work highlights the potential of active management. Preserving the forests that we have is more valuable than planting new forests. 
Pests and pathogens are driving mortality - may be opportunities for treatment/prevention of spread. 
Research into natural immunity (ash species in original range of EAB have immunity).
Management of deer populations to decrease browsing pressure - deer exclosure shows the importance to maintain an understory.

>>>>

Erika Gonzalez-Akre^1^  ( Orcid ID : 0000-0001-8305-6672)

@@

Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes. [@lovett_nonnative_2016; @herms_emerald_2014]. Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006]. Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018]. Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.

@@

In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig. 1).  The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig. 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years. This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013]. The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig. 1b). The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021. We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008). The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021].

@@

We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* [@gonzalez-akre_allodb_2022]. The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information. Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot study regions and the entire 25.6-ha plot. We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region. We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data (excluding stems < 10 cm dbh).

In addition to the estimation of AWR biomass flux, we analyzed the effects of nuisance species on recruitment through counts of newly recruited stems. We categorized each species as either canopy or understory based on expert knowledge of life history traits. *Quercus* spp. and *Carya* spp. were grouped by genera to better assess differences between the recruitment of canopy and understory species. We then calculated the number of new stems for each census interval for each species or genera. Finally, we assessed the spatial overlap of biomass loss and recruitment of canopy species. We calculated the ∆AGB of canopy trees within each 0.4 hectare quadrat between 2008 and 2023 and the AGB of canopy species in the understory in 2023.

@@

For the first time since the establishment of  the SCBI ForestGEO plot, living aboveground C stocks decreased over the 2018-2023 census interval (Figure 2), with AGB declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^. Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a). In regions characterized by high deer density and high canopy vulnerability, AGB decreased significantly over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023. Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.

![**_Figure `r biomass_trends`. Aboveground biomass at the SCBI ForestGEO plot since 2008 establishment._** Values are shown for the plot as a whole and for regions differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure2.jpeg)

 The transition from net AGB accumulation to loss is primarily attributable to increased AWM across all plot regions (Figure 3b). Initially, AWM flux remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first census intervals. However, in the most recent period, AWM increased significantly to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^. Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval. Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period. Simultaneously, the rate of biomass accrual from woody growth decreased across the plot (Figure 3c), declining from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over the 2008-2013 period to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval. Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the region with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^). Furthermore, AWG declined across all census intervals, even when AWM remained stable. Recruitment, constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure 3d). AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.

![**_Figure `r loss_gain_net`. Changes in aboveground woody biomass at the SCBI ForestGEO plot over each census interval since 2008 establishment, including (a) net change, (b) losses to mortality, and gains from (c) woody productivity and (d) recruitment._** Values are shown for the plot as a whole and for regions differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure3.jpeg)

@@

Over the 15-year study period, AWM increased throughout the plot yet AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure 4). *Fraxinus americana* had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhitibed significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.

![**_Figure `r mortality_by_genera`. Annual woody mortality by species at the SCBI ForestGEO plot from 2008-2023._** Shown are the six canopy taxa with highest average mortality rate in descending order and a category including all others. The census included all stems ≥ 10 cm in the most recent full census.](display/Figure4.jpeg)

## Species composition of recruitment

We report strong mediating effects of deer density on the quantity and composition of woody stem recruitment (Figure 5). The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^. In particular, the recruitment rate of *Carya* spp. and *Quercus* spp., late-successional taxa characteristic of this forest's expected climax state, was reduced by > 50% in areas with high deer density. The prevalence of vulnerable canopy species in regions with low deer density showed no significant impact on the total quantity of canopy species recruitment into the understory. However, recruitment of *Fraxinus americana* and *Fagus grandifolia* was higher in areas with high canopy vulnerability, while recruitment of *Ulmus rubra* and *Carya* spp. was lower. Across all regions, the recruitment rate of understory taxa was nearly 10 times that of canopy taxa. Recruitment of *Lindera benzoin* was high across all regions (> 100 stems ha^-1^ yr^-1^), and significantly increased in areas with high deer densities while peaking in areas that also had low canopy vulnerability. *Asimina triloba* recruitment was also lowest within the deer exclosure (`r round(group1_astrRecruit,1)` stems ha^-1^ yr^-1^) but was highest in the region with high deer density and high canopy vulnerability (`r round(group3_astrRecruit,1)` stems ha^-1^ yr^-1^).

![**_Figure `r recruitment_by_genera`. Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across areas differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment > 20x that of canopy species recruitment.](display/Figure5.jpeg)

## Overlapping biomass loss and recruitment

Our analyses reveal significant areas of decreasing AGB and suppressed regeneration of canopy species. Specifically, we found that 86% of the plot has either experienced -∆AGB from 2008-2023 or has a low AGB of canopy taxa in the understory. Only 21.7% of the plot exhibits an understory with high AGB of species capable of reaching the canopy. Within the region characterized by high deer density and high canopy vulnerability 84% of the area has lost canopy biomass, only 5.4% has an understory containing high AGB of canopy taxa, and 0% of area exhibits both. In contrast, 72.7% of the region with low deer density and low canopy vulnerability had a net increase of AGB. Furthermore, 70.7% of this same region has high AGB of canopy species in the understory, indicating a higher likelihood of future C sequestration.

![**_Figure `r regen_failure`. Characterizing the SCBI ForestGEO plot by changes in canopy biomass (2008-2023) and the biomass of canopy taxa in the understory in 2023._** (a) Each .04 ha quadrat is classified by the ∆AGB of canopy trees over the 15-year study period and the AGB of canopy taxa in the understory in 2023. (b) The % area of each region that falls within each classification.](display/Figure6.jpeg)

@@

Our investigation supports our hypotheses and demonstrates that the combination of -∆AGB and limited recruitment of canopy species in the understory is compromising the potential carbon sink of this forest, now and into the future. Decreases in AGB were largest in the regions of the plot with high canopy vulnerability to pests and pathogens. These regions also exhibited the lowest recruitment of canopy species in the understory. The region with low deer density had the highest amount of canopy species in the understory. Despite localized increases in AGB, notably in areas with low canopy vulnerability, on net, these increases are not outpacing AGB lost through a combination of increased AWM and decreasing AWG. Moreover, the low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.

We find that while nuisance species have had a large impact on AGB loss at our site (primarily through increased AWM of *Fraxinus americana*; Anderson-Teixeira et al, 2021), mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well. Oak species in the region are generally found in older, even aged cohorts, which can contribute to accelerated mortality rates (Oak et al., 2016). However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century. These mortality events are often termed “oak decline” events due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992). Oak decline is believed to be the result of an imbalance in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).

Relying on the current understory as a proxy for future canopy presents challenges. White-tailed deer populations have varied greatly over the last century (citation). Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events (Koenig, 2021). A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse, could allow for substantial, rapid regeneration of canopy species. However, current recruitment levels are below a rate which we would expect in maturing secondary forests. When overlaid with increasing mortality, regeneration may be insufficient to prevent extended periods without a closed canopy.

Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting broader applicability. Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests. Furthermore,the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context. EAB consistently induces near-total mortality in *Fraxinus* spp. once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).

Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forest ecosystem processes. Nuisance species, driven by human introduction and perturbations to ecological equilibriums, interact with other anthropogenic changes to modify forest ecosystems. By increasing mortality, nuisance species have the capacity to alter short-term carbon dynamics, resulting in large carbon losses that would not be expected in models that exclude them. Through recruitment suppression, nuisance species can have long-term effects that may alter successional dynamics. They can delay recovery from disturbances, anthropogenic or natural, and even suppress regeneration into a closed-canopy for sustained periods. They also can decrease resilience to other pressures as pests and pathogens decrease canopy diversity by shrinking the number of possible canopy species.

Nuisance species pose a serious threat to the health of temperate forests. But, this work highlights the potential of active management to protect them. Increased understory recruitment within the deer exclosure demonstrates that lowering deer populations can reduce browsing pressure. The direct linkage between non-indigenous pests and pathogens and decreased carbon storage demonstrates the need for natural climate solutions that focus on preserving the forests that we have instead of planting new ones. Opportunities exist to expand efforts that prevent further spread, increase the application of extant treatments, and continue research into natural immunity.
